### 2023-06-16 bachelorthesis/02-acknowledgement.tex
#39: Correction Florian Rath

I thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and brief communication, which helps me get better at English spelling. \bigskip

Outside of academia, I would like to extend my gratitude to my parents, my brother, his wife and daughter, my sister and my girlfriend for their encouragement, endurance and support while writing this thesis. I am aware that sometimes I am a workaholic who needs to be reminded about the life outside of my academic work.

>>>>

I thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and brief communication, which helps me getting better at English spelling. \bigskip

Outside of academia, I would like to extend my gratitude to my parents, my brother, his wife and daughter, my sister and my girlfriend for their encouragement, endurance and support while writing this thesis. 
% I am aware that sometimes I am a workaholic who needs to be reminded about the life outside of my academic work.

### 2023-06-16 bachelorthesis/21-motion.tex
#39: Correction Florian Rath

Due to the Lorentz force, particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field. This type of motion has circular frequency, which is often referred to as \textit{cyclotron frequency} and is defined as
\begin{gather}
    \omega_\mathrm{c} = \frac{|q|B}{m}~,
\end{gather}
where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by
\begin{gather}
    \rho_\mathrm{L} = \frac{mv_{\perp}}{|q|B}
\end{gather}
with the center often being referred to as \textit{gyrocenter}. Note that since the Lorentz force depends on the species of the particle, the circulation direction is the opposite between electron in ions.\\\bigskip

@@

In absence of forces in the parallel direction of the magnetic field the particles can move in parallel direction to the homogenous magnetic field. The velocity of this motion is connected to the thermal velocity $\vth$ and is dominated by electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).

@@

Since magnetic fields are not always homogenous, an inhomogeneous magnetic field with its gradient $\nabla B$ containing a component parallel to the magnetic field which is given by
\begin{gather}
    \nabla_{\!\parallel} B = \frac{\vect{B}}{B} \cdot \nabla B 
\end{gather}
causes the force
\begin{gather}
    F_{\parallel,\nabla_{\!\parallel} B} = - \frac{mv^2_{\perp}}{2B} \nabla_{\!\parallel} B = \mu \nabla_{\!\parallel} B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
\end{gather}
with \textit{magnetic moment} $\mu$. The magnetic moment $\mu$ can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$. The resulting force has its application in the mirror effect where a charged particle gets reflected due to this force. \cite{Wesson2011}

@@

\begin{enumerate}
    \item {\boldmath $\exb$} \textbf{Drift:}\\
    If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results to a drift of the gyrocenter with
    \begin{gather}
        \vect{v}_{E} = \frac{\vect{E}\times\vect{B}}{B^2}
        \label{eq:exb}
    \end{gather}
    which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
    \item {\boldmath $\nabla B$} \textbf{Drift:}\\
    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$ drift defined by
    \begin{gather}
        \vect{v}_{\nabla B} = \frac{m v^2_{\perp}}{2 q}\frac{\vect{B}\times \nabla B}{B^3}~.
        \label{eq:gradB}
    \end{gather}
    The $\nabla B$ drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type.
    \item \textbf{Curvature Drift:}\\
    Due to centrifugal force acting on the particle in a curved magnetic field the gyrocenter experiences a curvature drift according to
    \begin{gather}
        \vect{v}_{\kappa} = \frac{m v^2_{\parallel}}{q}\frac{\vect{B}\times \vect{\kappa}}{B^2} = \frac{m v^2_{\parallel}}{q} \frac{\vect{B}\times \nabla B}{B^3}~;\qquad\vect{\kappa} = -(\vect{b}\cdot \nabla)\vect{b} = \frac{\nabla B}{B}~,
        \label{eq:curvature}
    \end{gather}
    where $\vect{b}$ is the unit vector along the magnetic field. To obtain the result for the curvature $\vect{\kappa}$ in Eq. (\ref{eq:curvature}) the plasma pressure has to be small compared to the magnetic field strength $B$. In the form of Eq. (\ref{eq:curvature}) $\nabla B$ and curvature drift can be treated similarly. \cite{Wesson2011}
\end{enumerate}

>>>>

Due to the Lorentz force, particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field. This type of motion has circular frequency, which is often referred to as \textit{cyclotron frequency} and is defined as
\begin{gather}
    \omega_\mathrm{c} = \frac{|q|B}{m}~,
\end{gather}
where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by
\begin{gather}
    \rho_\mathrm{L} = \frac{mv_{\perp}}{|q|B}
\end{gather}
with the center often being referred to as \textit{gyrocenter}. Note that since the Lorentz force depends on the species charge of the particle, the circulation direction is the opposite between electron in ions.\\\bigskip

@@

In absence of forces in the direction parallel to the magnetic field the particles can move freely in parallel direction to the homogenous magnetic field. The velocity of this motion is of order of the thermal velocity $\vth$ and is dominated by electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).

@@

Since magnetic fields are not always homogenous, an inhomogeneous magnetic field with its gradient $\nabla B$ containing a component parallel to the magnetic field which is given by
\begin{gather}
    \nabla_{\!\parallel} B = \frac{\vect{B}}{B} \cdot \nabla B 
\end{gather}
causes the force
\begin{gather}
    F_{\parallel,\nabla_{\!\parallel} B} = - \frac{mv^2_{\perp}}{2B} \nabla_{\!\parallel} B = \mu \nabla_{\!\parallel} B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
\end{gather}
with \textit{magnetic moment} $\mu$. The magnetic moment $\mu$ is an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$. The resulting force has its application in the mirror effect where a charged particle gets reflected due to this force. \cite{Wesson2011}

@@

\begin{enumerate}
    \item {\boldmath $\exb$} \textbf{Drift:}\\
    If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results into a drift of the gyrocenter with
    \begin{gather}
        \vect{v}_{E} = \frac{\vect{E}\times\vect{B}}{B^2}
        \label{eq:exb}
    \end{gather}
    which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
    \item {\boldmath $\nabla B$} \textbf{Drift:}\\
    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$ drift defined by
    \begin{gather}
        \vect{v}_{\nabla B} = \frac{m v^2_{\perp}}{2 q}\frac{\vect{B}\times \nabla B}{B^3}~.
        \label{eq:gradB}
    \end{gather}
    The gradient of the magnetic field $\nabla B$ varies thereby on scales larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type.
    \item \textbf{Curvature Drift:}\\
    Due to centrifugal force acting on the particle in a curved magnetic field the gyrocenter experiences a curvature drift according to
    \begin{gather}
        \vect{v}_{\kappa} = \frac{m v^2_{\parallel}}{q}\frac{\vect{B}\times \vect{\kappa}}{B^2} = \frac{m v^2_{\parallel}}{q} \frac{\vect{B}\times \nabla B}{B^3}~;\qquad\vect{\kappa} = -(\vect{b}\cdot \nabla)\vect{b} = \frac{\nabla B}{B}~,
        \label{eq:curvature}
    \end{gather}
    where $\vect{b}$ is the unit vector along the magnetic field. To obtain the result for the curvature $\vect{\kappa}$ in Eq. (\ref{eq:curvature}) the plasma pressure has to be small compared to the magnetic field strength $B$. In the form of Eq. (\ref{eq:curvature}) $\nabla B$ and curvature drift can be treated similarly. \cite{Wesson2011}
\end{enumerate}

### 2023-06-16 bachelorthesis/22-confinement.tex
#39: Correction Florian Rath

In tokamak devices strong magnetic fields confine the hot plasma. As mentioned in Chapter \ref{sec:motion} a magnetic field forces a perpendicular particle motion and a motion which contains the gyro motion and slow perpendicular gyro center drifts. Because of the much smaller size of the Larmor radius compared to the device size $R$ the particle and energy losses are caused by the gyro center drift. To avoid this at the end of the field lines the magnetic field in the tokamak devices is shaped like a torus. This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux surfaces}, and magnetic field lines which appear on these surfaces. To maintain stability the magnetic field has a toroidal and a poloidal component which also is equivalant to the plasma pressure. \cite{Stroth2011, Wesson2011} The toroidal component is produced by external coils whereas the poloidal component is provided by the toroidal plasma current. Together the components result in a magnetic field which follows helical trajectories [Fig \ref{fig:confinement}]. To characterize the quality of confinement the so-called \textit{plasma beta} is used and is given as
\begin{gather}
    \beta = \frac{nT}{\mu_0 B^2/2}~,
\end{gather} 
with $n$ the plasma density, $T$ as temperature, $\mu_0$ the permeability in vacuum and the magnetic field strength $B$. Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$ which indicates that in cases when the plasma beta is smaller the confinement is better. For fusion devices the plasma beta has to be smaller than 1 ($\beta < 1$). In a tokamak reactor the plasma beta has a typical order of a few percent. \cite{Wesson2011}
\includegraphicsHere{Theory/Tokamak-Torus.pdf}{
	Toroidal flux surfaces in tokamak plasma with helical magnetic field (\textcolor{Ubtgreen}{green} line) in torus coordinates ($\rho$ (radial), $\phi$ (toroidal), $\theta$ (poloidal)) or cylindrical coordinates ($Z$, $R$, $\phi$). \cite{Barton2015}
}{fig:confinement}{0.7}

>>>>

In tokamak devices strong magnetic fields confine the hot plasma. As mentioned in Chapter \ref{sec:motion} a magnetic field forces a perpendicular particle motion and a motion which contains the gyro motion and slow perpendicular gyro center drifts. Because of the much smaller size of the Larmor radius compared to the device size $R$ the particle and energy losses are caused by the gyro center drift. To avoid additional loss of particles because of the parallel motion the field lines of the magnetic field in the tokamak devices is shaped like a torus. This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux surfaces}, and magnetic field lines which lie on these surfaces. To maintain stability the magnetic field has a toroidal and a poloidal component. According to the force balance the magnetic field is equivalent to the plasma pressure which means on flux-surfaces the plasma pressure is constant. \cite{Stroth2011, Wesson2011} The toroidal component is produced by external coils whereas the poloidal component is provided by the toroidal plasma current. Together the components result in a magnetic field which follows helical trajectories [Fig \ref{fig:confinement}]. To characterize the quality of confinement the so-called \textit{plasma beta} is used and is given as
\begin{gather}
    \beta = \frac{nT}{\mu_0 B^2/2}~,
\end{gather} 
with $n$ the plasma density, $T$ as temperature, $\mu_0$ the permeability in vacuum and the magnetic field strength $B$. Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$. For fusion devices the plasma beta has to be a bit smaller than 1 ($\beta < 1$) for optimal confinement. In a tokamak reactor the plasma beta has a typical order of a few percent. \cite{Wesson2011}
\includegraphicsHere{Theory/Tokamak-Torus.pdf}{
	Toroidal flux surfaces in tokamak plasma with helical magnetic field (\textcolor{Ubtgreen}{green} line) in torus coordinates ($\rho$ (radial), $\phi$ (toroidal), $\theta$ (poloidal)) or cylindrical coordinates ($Z$, $R$, $\phi$). \cite{Barton2015}
}{fig:confinement}{0.7}

### 2023-06-16 bachelorthesis/23-iontempgrad.tex
#39: Correction Florian Rath

The described electrostatic forces acting on the tokamak plasma cause a variety of collective phenomena. The occurence of Microinstabilities is one of these phenomena. They are instabilities on the spatial scale of the Larmor radius $\rho_\mathrm{L}$ and frequencies much smaller than the cyclotron frequency $\omega_\mathrm{c}$, driven by density and temperature gradients present in the tokamak devices. Turbulence by microinstabilities are considered as the main reason for the loss of particle energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors. The most dominant instability in tokamak plasma is the ion temperature gradient (ITG) driven instability. \cite{Coppi1967, Cowley1991, Rudakov1961}

@@

The ITG driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak. The mechanism is briefly explained with the help of Fig. \ref{fig:ITG} and Ref. \citenum{Beer_PHD, Casson_PHD, Dannert_PHD}. An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature drift and $\nabla B$ drift with velocity $\vect{v}_\mathrm{d}$, which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells). This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response. As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$. The $\exb$ drift direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip

The described mechanism causes perturbation on the outboard low field side of the tokamak this region has a \enquote{bad curvature} and instability happens there. On the inboard high field side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}. Furthermore, such turbulent structures are called \textit{eddies} which have a typical radial scale of serval Larmor radius. \cite{Newins2006}
The initial perturbation is growing if the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$. For nonlinear turbulence the critical value is given by $R/L_\mathrm{T,c} \sim 6$. \cite{Dimits2000, Isliker2010}

>>>>

The described electrostatic forces acting on the tokamak plasma cause a variety of collective phenomena. The occurrence of microinstabilities is one of these phenomena. They are instabilities on the spatial scale of the Larmor radius $\rho_\mathrm{L}$ and frequencies much smaller than the cyclotron frequency $\omega_\mathrm{c}$, driven by density and temperature gradients present in the tokamak devices. Turbulence by microinstabilities are considered as one of the main reason for the loss of particle energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors. The most dominant instability in tokamak plasma is the ion temperature gradient (ITG) driven instability. \cite{Coppi1967, Cowley1991, Rudakov1961}

@@

The ITG driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak. The mechanism is briefly explained with the help of Fig. \ref{fig:ITG} and Ref. \citenum{Beer_PHD, Casson_PHD, Dannert_PHD}. An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the modulation of curvature drift and $\nabla B$ drift with velocity $\vect{v}_\mathrm{d}$. The divergence of the velocity $\vect{v}_\mathrm{d}$ is therefore not zero which results in compression of the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells). This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response. As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$. The $\exb$ drift direction transports hot plasma in the outer hot region of perturbation ($T^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($T^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip

The described mechanism causes perturbation on the outboard low field side of the tokamak this region has a \enquote{bad curvature} and instability happens there. On the inboard high field side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}. Furthermore, such turbulent structures are called \textit{eddies} which have a typical radial scale of several Larmor radii. \cite{Newins2006}
The initial perturbation is growing if the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$. For nonlinear turbulence with Cyclone Base Case parameters and circular concentric flux-surfaces the critical value is given by $R/L_\mathrm{T,c} \sim 6$. \cite{Dimits2000, Isliker2010}

### 2023-06-16 bachelorthesis/24-zonalflows.tex
#39: Update headings with bold math
#39: Correction Florian Rath

\section{Zonal Flows and Shearing Rate {\boldmath $\wexb$}}
\label{sec:zonalflow}

Zonal flows are linked in the tokamak plasma to the $\exb$ flows [Eq.(\ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}. \cite{Diamond2005} Because of the variation exclusively across flux-surfaces, the zonal flow is tangential to those surfaces and does not contribute to the turbulent transport across the flux-surfaces caused by the $\exb$ motion. 
Zonal flows are driven nonlinearly by turbulent Reynolds stresses. \cite{Diamond1991} For more details about the mechanism behind zonal flow generation the reader is referred to Ref. \citenum{Brizard2007, Dubin1983, Hasegawa1978, Krommes2000}. \bigskip

@@

Furthermore, zonal flows are connected with the radially alternating zonal potential, which introduces a velocity shear, i.e., a radial variation of the zonal flow velocity. This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000}, especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}. The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a later development of ITG driven turbulence. The shearing process tilts and stretches turbulent eddies, caused by the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig. \ref{fig:zonalflow}],  characterizing the shearing as a decorrelation process. Turbulence will then be reduced because of this process. \cite{Biglari1990,Diamond2005,Burnell1997} \bigskip

@@

As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation. Avalanches are initiated when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards with positive shearing. \cite{Idomura2009,McMillan2009} The fully developed avalanche results in almost completely quenched turbulence. Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold}. \cite{Peeters2016,Weikl2017}

>>>>

\section{Zonal Flows and Shearing Rate $\wexb$}
\label{sec:zonalflow}

In the tokamak plasma zonal flows are linked to the $\exb$ flows [Eq.(\ref{eq:exb})] that are connected to the electrostatic potential that is constant on flux-surfaces, also called \textit{zonal potential}. \cite{Diamond2005} Because of the variation exclusively across flux-surfaces, the zonal flow is tangential to those surfaces and does not contribute to the turbulent transport across the flux-surfaces caused by the $\exb$ motion. 
Zonal flows are driven nonlinearly by turbulent Reynolds stresses. \cite{Diamond1991} For more details about the mechanism behind zonal flow generation the reader is referred to Ref. \citenum{Brizard2007, Dubin1983, Hasegawa1978, Krommes2000}. \bigskip

@@

Furthermore, zonal flows are connected with the radially alternating zonal potential, which introduces a velocity shear, i.e., a radial variation of the zonal flow velocity. This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000}, especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}. The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a development of ITG driven turbulence at higher $\rlt$ compared to the linear threshold. The shearing process tilts and stretches turbulent eddies, caused by the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig. \ref{fig:zonalflow}],  characterizing the shearing as a decorrelation process. Turbulence will then be reduced because of this process. \cite{Biglari1990,Diamond2005,Burnell1997} \bigskip

@@

As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation. Avalanches are initiated when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards with positive shearing. \cite{Idomura2009,McMillan2009} The fully developed staircase results in almost completely quenched turbulence. Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient also known as \textit{finite heat flux threshold}. \cite{Peeters2016,Weikl2017}
